\NAME: A framework for \\ Efficient Intrusion Detection\\ in the Internet of Things

### 2014-11-21 paper.tex
processed Danny’s remarks on body

The internet of things (IoT) holds a great potential to positively influence
our daily work and life through domotics, assisted living, e-health and
enhanced learning \cite{atzori2010internet}. But with this potential the IoT
also presents a significant threat: by opening our smart homes and personal
data to all these interconnected devices, we also open them to everyone who is
able to break the virtual locks that protect them (i.e.\ passwords, encryption
keys,\dots). IoT devices typically have limited batteries, processing power and
memory. These properties make it hard to add any additional service, including
security.

@@

In classic networks, firewalls focus on the outer perimeter of the network,
filtering unwanted packets and protecting the entire internal network. But
attacks on flaws in services can pass unnoticed. This is where intrusion
detection (ID) comes into play: an intrusion detection system (IDS) monitors
all traffic that passes through the firewall, looking for patterns of malicious
activity, and optionally, after detecting such pattern, alerting the firewall,
allowing it to take corrective actions \cite{denning1987intrusion}.

@@

In wireless networks of resource-constraint devices, which make up a
significant part of the IoT, it is not possible to have this single point in
the network that can oversee all traffic \cite{mishra2004intrusion}. Every
device has to implement its own lines of defense. But making the IDS a local
service on each device requires local resources, which are a scarce commodity
for IoT devices.

Attackers have access to a large set of possible attack vectors
\cite{aschenbruck2012security}, ranging from the wireless network, over the
medium access control and routing, up to the device itself. Each level presents
different opportunities to manipulate data, eavesdrop, perform a form of denial
of service or even access the network in a legitimate way. For each of these
attacks an algorithm needs to look for patterns or anomalies. An IDS for IoT
devices therefore requires a large number of implemented algorithms for an
adequate coverage.

@@

Current developments in ID focus on programming frameworks to structure the
implementation of algorithms \cite{valero2012di} and offer the required basic
functional components to implement them \cite{krontiris2008lidea}. The first
aspect that is missing, is that they don't offer a way to optimize the usage of
resources. A solution that supports the integration of multiple algorithms on
IoT devices needs to avoid accumulating their impact on the available resources.

A second omission is support for systematic reuse to create different
configurations for different devices, both in hard- and software. Addressing
this in a transparent and automated way is important for any kind of
development in this area.

@@

The second contribution introduces functional code fusion (FCF) as a source
code generation paradigm to solve the problem of hard to reuse and combine ID
algorithms. FCF identifies common data and functions, and organizes code to
eliminate redundant iterations, tests and computations. Results show improved
image size, usage of the wireless radio, execution time and memory usage.

@@

Research into ID in wireless networks typically focuses on two major topics:
the devices as single entities and the network as a group of such devices. Both
are important: a group cannot make a decision without members that detect
malicious behavior and a group-based decision is often needed because devices
in the network can miss certain events that could reveal intrusions. In a
wireless network, where not all participants are within each others' radio
range, wireless devices typically only have a partial view on the entire
network. This prohibits them to observe potentially crucial events in distant
parts of the network. The fact that resource-constraint devices typically apply
duty cycling to lower their energy consumption, further increases this problem,
introducing periods of time where they even don't observe local events.

@@

  \item[Broadcast communication] can allow implementing a distributed,
  cooperative algorithm locally. The reputation-based detection algorithm
  introduced in \cite{ganeriwal2008reputation} illustrates this. It uses the
  same principle of a watchdog also found in \cite{mishra2004intrusion}. Using
  this algorithm, devices can exchange information about the reputation of
  other devices, solely using one-way broadcast messages. Combining this
  information allows them to decide about their trust in a given device, based
  on more than their own, often partial, observations.

@@

Given this classification, we now introduce a design pattern that identifies
common functional aspects of these algorithms. It will allow us to define a
strategy for optimizing the combination of these algorithms. From the
classification and a survey of algorithms, we extract two activities common to
most algorithms: the inspection of network traffic and the evaluation of
thresholds.

@@

But simply checking a threshold after updating some counter, is not a good
strategy. A simple example is that of a watchdog \cite{mishra2004intrusion},
observing the actions of other devices. There are valid reasons why a device
can miss a theoretical window in which one expects activity, i.e.\ the wireless
network being the most predominantly unreliable component. Directly reacting on
a basic threshold can result in false positives. A way to overcome this is to
decouple the detection of problems and the decision what to do about them.

@@

\begin{figure}[ht]
\captionof{lstlisting}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
\begin{algorithmic}[1]
  \Require{devices: global storage for device information}
  \Function{process\_message}{$msg$}
    \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
     \State \dots \Comment{analyze byte-sequences}
    \EndFor
    \Let{$devices_x$}{value}  \Comment{optionally update device info}
    \State \Call{send}{$devices_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
  \EndFunction
  \State
  \Function{do\_housekeeping}{}
    \ForEach{$device \in devices$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
      \If{$device > \dots$} \Comment{validate recorded value}
        \State \dots \Comment{take actions}
        \State \Call{send}{$device$, ``info''} \Comment{communicate} \label{alg:id-algo-pattern-send2}
      \EndIf
    \EndFor
  \EndFunction
  \State
\end{algorithmic}
\end{figure}

@@

  \item[Loops] by themselves are not a problem, but different loops, that
  perform the same operations in sequence, can incur an overhead. An example
  from the ID algorithm pattern is the handling of messages: each algorithm
  parses each observed message on the network, processing the same byte stream,
  performing comparable operations to find similar patterns.

@@

  \item[The heterogeneity of IoT devices] results in different software stacks,
  each with for example their own network API\@. This also explains why there
  are no implementations readily available to developers: there is no single
  platform to target for researchers.

@@

Sending and receiving of network messages is tightly integrated with the
previously mentioned concepts of event handling and the Nodes domain. An
underlying software library can handle all interaction with the network,
triggering functions by means of events when parsing new messages and exposing
an API to send messages through the Nodes domain. This approach gives the
framework complete control over communication, allowing optimized marshaling,
grouping,\dots

@@

Listing \ref{lst:watchdog} illustrates some of the features mentioned above,
using a small but representative example of an algorithm: a
watchdog \cite{mishra2004intrusion}. Like a heartbeat, nodes can broadcast a
message at a regular interval, allowing other nodes to monitor its
availability. Missing heartbeats could indicate problems, such as physical
capturing of a device.

@@

Relying on basic hardware and software allows us to validate that the generator
is capable of generating code even for the most basic environment. Any step up,
both in hardware or software, would offer better and higher abstractions that
would make it easier to generate code for that situation. It shows that the
requirements of the generator towards the platform are minimal, don't rely on
advanced frameworks or operating systems and therefore are applicable in any
environment, targeting any platform.

@@

Most remarkable is the effect on the time of one event loop cycle: both
algorithms by itself add 152\%. Adding a second algorithm only augments the
impact to 188\%. We see the effect of the generic handling of common
functionality, like message parsing, event scheduling,\dots It comes with an
overhead for a single algorithm, but pays for itself when adding more
algorithms.

@@

Secondly, we see that the software library that comes with the generated code
adds to the size of the resulting image. But the cost is close to constant and
is lower for both algorithms together. With roughly 3KB of additional overhead,
or 15\% in this simple case, the impact is small. Taking into account the
available memory on the ATMega1284p, this is merely 2.3\% of the available
resources.

Thirdly, when turning to the goal to reduce the network usage, we see that both
the number of packets as well as the amount of bytes sent have dropped by 13
and 7\%.

@@

SEAL \cite{elsts2013seal} is a novel complete programming language that focuses
on novice programmers. It is a clear fit for \emph{sense-and-send} type of
applications, and doesn't pretend to be the best choice for any type of
application. It does however share the idea of event to overcome the technical
nature of loops.

@@

Finally, applying code generation to IDS, typically tries to generate native
code to avoid the interpretation of the description of patterns.
I2S \cite{charitakis2003code} is an example that generates code for
SNORT \cite{roesch1999snort} rules. To optimize this highly efficient code, it
often targets a specific platform.

@@

Adequate intrusion detection for IoT devices requires multiple algorithms to
operate at the same time. Recent literature describes different algorithms, but
hardly any are readily available for direct application when implementing an
IDS for IoT devices. Reselection of algorithms, due to changing requirements or
security threats, requires a flexible way to incorporate the different
algorithms. This eliminates manual implementation and optimization.

Wireless resource-constraint devices, such as IoT devices, require that
software respects their restrictions. Adding multiple intrusion detection
algorithm requires fusion of these algorithms on a functional level, to reduce
execution inefficiencies, redundant memory usage and superfluous network access.

\NAME is a framework consisting of a DSL, a source code generator and
accompanying software library. Using the DSL, intrusion detection algorithms
can be formally described in terms of functionality. During source code
generation, functional code fusion interprets the functional meaning of the
algorithms and merges common functionality and shared data. The software
library hosts common functionality, such as network message parsing, sent
message aggregation, event scheduling,\dots This allows the source code
generator to focus on the optimization of the organization of the functionality.

>>>>

The internet of things (IoT) holds great potential to positively influence our
daily work and life through domotics, assisted living, e-health and enhanced
learning \cite{atzori2010internet}. But with this potential the IoT also
presents a significant threat: by opening our smart homes and personal data to
all of these interconnected devices, we also open them to everyone who is able
to break the virtual locks that protect them (i.e.\ passwords, encryption
keys). IoT devices typically have limited batteries, processing power and
memory. These properties make it hard to add any additional service, including
security.

@@

In classic networks, firewalls focus on the outer perimeter of the network,
filtering unwanted packets and protecting the entire internal network. But
attacks on flaws in services can pass unnoticed. This is where intrusion
detection (ID) comes into play: an intrusion detection system (IDS) monitors
all traffic that passes through the firewall, looking for patterns of malicious
activity, and optionally, after detecting such pattern, alerts the firewall,
allowing it to take corrective actions \cite{denning1987intrusion}.

@@

In wireless networks of resource-constrained devices, which make up a
significant part of the IoT, it is not possible for a single point in the
network to oversee all traffic \cite{mishra2004intrusion}. Every device has to
implement its own lines of defense. But making the IDS a local service on each
device requires local resources, which are a scarce commodity for IoT devices.

Attackers have access to a large set of possible attack vectors
\cite{aschenbruck2012security}, ranging from the wireless network, through the
medium access control and routing, up to the device itself. Each level presents
different opportunities to manipulate data, eavesdrop, perform a form of denial
of service or even access the network in a legitimate way. For each of these
attacks an algorithm needs to look for patterns or anomalies. An IDS for IoT
devices therefore requires a large number of implemented algorithms for an
adequate coverage.

@@

Current developments in ID focus on programming frameworks that structure the
implementation of algorithms \cite{valero2012di} and offer the required basic
functional components to implement them \cite{krontiris2008lidea}. A key
shortcoming of prior ID frameworks, is that they don't offer a way to optimize
the usage of resources. A solution that supports the integration of multiple
algorithms on IoT devices needs to avoid accumulating their impact on the
available resources.

A second omission is support for systematic reuse to create different
configurations heterogeneous devices. Addressing this in a transparent and
automated way offers the potential to significantly reduce the effort of
developing an IDS.

@@

The second contribution introduces functional code fusion (FCF) as a source
code generation paradigm to solve the problem of hard to reuse and combine ID
algorithms. FCF identifies common data and functions, and organizes code to
eliminate redundant iterations, tests and computations. Results show improved
image size, reduced usage of the wireless radio, reduced execution time and
lower memory usage.

@@

Research into ID in wireless networks typically focuses on two major topics:
the devices as single entities and the network as a group of such devices. Both
are important: a group cannot make a decision without members that detect
malicious behavior and a group-based decision is often needed because devices
in the network can miss certain events that could reveal intrusions. In a
wireless network, where not all participants are within each others' radio
range, wireless devices typically only have a partial view on the entire
network. This prohibits them from observing potentially crucial events in
distant parts of the network. The fact that resource-constrained devices
typically apply duty cycling to lower their energy consumption, further
increases this problem, introducing periods of time where they cannot observe
local events.

@@

  \item[Broadcast communication] can allow implementing a distributed,
  cooperative algorithm locally: devices can exchange information about the
  reputation of other devices, solely using one-way broadcast messages.
  Combining this information allows them to decide about their trust in a given
  device, based on more than their own, often partial, observations. The
  reputation-based detection algorithm introduced in
  \cite{ganeriwal2008reputation} illustrates this. It uses the same principle
  of a watchdog also found in \cite{mishra2004intrusion}.

@@

Based upon the classification presented in section \ref{classification}, we now
introduce a design pattern that identifies common functional aspects of ID
algorithms. It will allow us to define a strategy for optimizing the
combination of these algorithms. We extract two activities common to most
algorithms: the inspection of network traffic and the evaluation of thresholds.

@@

Updating such statistics and evaluating the corresponding thresholds should be
distinct operations. The example of a watchdog \cite{mishra2004intrusion}
illustrates this: a device monitors the actions of other devices and checks if
they correctly forward messages to distant nodes in the network. Missing
forwards can indicate a malicious intension to silently drop messages and
obstruct the normal operation of the network. There are valid reasons why a
device can miss the theoretical window in which one expects activity, i.e.\ the
wireless network being the most predominantly unreliable component. Directly
reacting upon a basic threshold can result in false positives. A way to
overcome this is to decouple the detection of problems and the decision what to
do about them.

@@

\begin{figure}[ht]
\captionof{lstlisting}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
\begin{algorithmic}[1]
  \Require{devices: global variable for device information}
  \Function{process\_message}{$msg$}
    \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
     \State \dots \Comment{analyze byte-sequences}
    \EndFor
    \Let{$devices_x$}{value}  \Comment{optionally update device info}
    \State \Call{send}{$devices_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
  \EndFunction
  \State
  \Function{do\_housekeeping}{}
    \ForEach{$device \in devices$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
      \If{$device > \dots$} \Comment{validate recorded value}
        \State \dots \Comment{take actions}
        \State \Call{send}{$device$, ``info''} \Comment{communicate} \label{alg:id-algo-pattern-send2}
      \EndIf
    \EndFor
  \EndFunction
  \State
\end{algorithmic}
\end{figure}

@@

  \item[Loops] by themselves are not a problem, but different loops, that
  perform the same operations in sequence, can incur overhead. An example from
  the ID algorithm pattern is the handling of messages: each algorithm parses
  each observed message on the network, processing the same byte stream,
  performing comparable operations to find similar patterns.

@@

  \item[The heterogeneity of IoT devices] results in different software stacks,
  each with for example their own network API\@. This also explains why there
  are few implementations readily available to developers: there is no single
  platform to target for researchers.

@@

Sending and receiving of network messages is tightly integrated with the
previously mentioned concepts of event handling and the Nodes domain. An
underlying software library can handle all interaction with the network,
triggering functions by means of events when parsing new messages and exposing
an API to send messages through the Nodes domain. This approach gives the
framework complete control over communication and for example allows for
optimized marshaling and grouping of messages.

@@

Listing \ref{lst:watchdog} illustrates some of the features mentioned above,
using a small but representative example of an algorithm: a watchdog
\cite{mishra2004intrusion}. Like a heartbeat, nodes can broadcast a message at
a regular interval, allowing other nodes to monitor its availability. Missing
heartbeats could indicate problems, such as physical capture of a device.

@@

Relying on basic hardware and software allows us to validate that the generator
is capable of generating code even for the most basic environment. It shows
that the requirements of the generator towards the platform are minimal, don't
rely on advanced frameworks or operating systems and therefore are applicable
in any environment, targeting any platform.

@@

Most remarkable is the effect on the time of one event loop cycle: both
algorithms by itself add 152\%. Adding a second algorithm only augments the
impact to 188\%. We see the effect of the generic handling of common
functionality, like message parsing and event scheduling. It comes with an
overhead for a single algorithm, but pays for itself when adding more
algorithms.

@@

Secondly, we see that the software library that comes with the generated code
adds to the size of the resulting image. But the cost is close to constant and
is lower for both algorithms together. With roughly 3KB of additional overhead,
or 15\% in this simple case, the impact is small. Taking into account the
available memory on the ATMega1284p, this is only 2.3\% of the available
resources.

Thirdly, when turning to the goal to reduce the network usage, we see that both
the number of packets as well as the amount of bytes sent have dropped by 13\%
and 7\% respectively.

@@

SEAL \cite{elsts2013seal} is a novel complete programming language that focuses
on novice programmers. It is a clear fit for \emph{sense-and-send} type of
applications. It does however share the idea of events to overcome the technical
nature of loops.

@@

Applying code generation to the context of ID, often focusses on the generation
of native code to avoid the interpretation of descriptions of patterns to look
for. I2S \cite{charitakis2003code} is an example that generates code for SNORT
\cite{roesch1999snort} rules. To optimize this highly efficient code, it often
targets a specific platform.

@@

Adequate intrusion detection for IoT devices requires multiple algorithms to
operate at the same time. Recent literature describes different algorithms, but
few are readily available for direct application when implementing an IDS for
IoT devices. Reselection of algorithms, due to changing requirements or
security threats, requires a flexible way to incorporate the different
algorithms. This eliminates manual implementation and optimization.

Wireless resource-constraint devices, such as IoT devices, require that
software respects their restrictions. Adding multiple intrusion detection
algorithms requires fusion of these algorithms on a functional level, to reduce
execution inefficiencies, redundant memory usage and superfluous network access.

\NAME is a framework consisting of a DSL, a source code generator and
accompanying software library. Using the DSL, intrusion detection algorithms
can be formally described in terms of functionality. During source code
generation, functional code fusion interprets the functional meaning of the
algorithms and merges common functionality and shared data. The software
library hosts common functionality, such as network message parsing, sent
message aggregation and event scheduling. This allows the source code generator
to focus on the optimization of the organization of the functionality.
